pollen_countR

### 2023-02-15 script/manuscript_MEE.Rmd
new results in, lower detection?

--- 
title: Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging Patterns of Queen Bumble Bees
author:  |
    |  Sophie Taddeo$^1$$^,$$^2$, Paul J. CaraDonna$^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant$^1$$^,$$^2$  
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    |  $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, Colorado 81224, USA
    |  $^4$ Australia post fire project, ----, ----, ----, AU  
    |  Reed Clark Benkendorf$^1$$^,$$^2$^[Correspondence: reedbenkendorf2021@u.northwestern.edu], Jane E. Ogilive$^3$$^,$$^4$, Emily J. Woodworth$^1$$^,$$^2$,
abstract:  |  
  |
  1) DNA Barcoding has been successful for the rapid analysis of ecological assemblages. Although barcoding in the plant kingdom barcoding has been more difficult than others, and hence may begin to lag behind other kingdoms. 
  2) Here we test the utilization of Angiosperms 353 probes to barcode plant species found in pollen loads collected from Queen Bumble Bees. 
  3) To verify the accuracy for this barcoding system we compared the data to museum species, observation studies, and species distribution modelling to identify likely candidate species. 
  4) By utilizing Species distribution modelling we allow users to create a regionally appropriate sequence databases which may use increase the alignment algorithms minimizing need for large computational power, and run time. 
  5) We show that the Angiosperms 353 probes, which are currently being used in the largest ever plant systematic endeavor, offers significant promise to metagenomic approaches. 
  6) Understanding plants in ecological contexts and understandings of their synecology.
keywords: |
  DNA baroding, metagenomics, hyb-seq, Angiosperms353, plant metagenomics, bombus queens, subalpine, species distribution model
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: true
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

pmorph <- read.csv('../data/morpho_pollen.csv') 
blst <- read.csv('../data/Post_Classified_BLAST_sqs.csv') %>% 
  rename(sample.id = Sample, Family = FAMILY, Genus = GENUS) %>% 
  group_by(sample.id) %>% 
  mutate(Prcnt_Seqs_Class = sum(Prcnt_seqs)) %>% 
  ungroup()
blst_seqs <- distinct(blst, sample.id)

@@

# PERFORM FOR ERICACEAE TOO - A single record maybe come out?
ericaceae_samples <- pmorph %>% 
  filter(morphotype == 'ERICACEAE') %>% 
  select(sample.id, Prop, morphotype)

blst_ericoid <- filter(blst, Family == 'Ericaceae') %>% # 44 might be Epilobium anagallidolium
  mutate(morphotype = 'ERICACEAE')
ericaceae_detected <- ericaceae_samples %>% 
  mutate(DETECTED = if_else(sample.id %in% blst_ericoid$sample.id, 1, 0))

@@

aa <- prcnt_detect(mertensia_detected, Lprop = 0.01, Hprop = 1)
bb <- prcnt_detect(aster1_detected, Lprop = 0.001, Hprop = 0.01)
cc <- prcnt_detect(aster2_detected, Lprop = 0.001, Hprop = 0.01)
dd <- prcnt_detect(aster_detected, Lprop = 0.01, Hprop = 1)
ee <- prcnt_detect(ericaceae_detected, Lprop = 0.001, Hprop = 0.1)
ff <- prcnt_detect(aster_detected, Lprop = 0.001, Hprop = 0.01)

# Semi-quantitative accord between sequence % and Grain Count
palyn_cnt <- bind_rows(aster1_samples, aster2_samples, ericaceae_samples, mertensia_samples) %>% 
  select(morphotype, countProp = Prop)

pollen_props <- bind_rows(blst_aster1, blst_aster2, blst_mertensia, blst_ericoid) %>% 
  select(morphotype, sequenceProp = Prcnt_seqs, Seqs_pr_taxon, Total_seqs, Prcnt_Seqs_Class) %>% 
  left_join(., palyn_cnt, by = c('sample.id', 'morphotype')) %>% 
  drop_na()

rm(mertensia_samples, blst_aster, blst_aster1,
   blst_aster2, aster1_samples,  aster2_samples, a, b, c, d, palyn_cnt,
   blst_mertensia, aster_samples, ericaceae_samples, blst_ericoid)
```

@@

knitr::kable(pp_seqs_t, align = "c", 
             col.names = c('Condition', 'No. Class.', 'Prcnt. Class.', 'Total Seqs', 'Rank'),
             caption = "Post classification of Sequences via Taxonomy and Ecology") %>% 
  kableExtra::kable_styling(
      latex_options = "striped",
      stripe_color = "gray!10")

@@

To determine at which level species in pollen loads could be detected the results of light microscopy were compared to the molecular results. The pollen samples contained three morphotypes which could readily be identified via microscopy. Two of these mapped to the clades (Boraginaceae & Heliantheae Alliance), and one to a Asteraceae less Heliantheae. Boraginaceae grains `r toString(aa)`.  Asteraceae type 1, non-helianthoids, `r toString(bb)` Asteraceae type 2, Helianthoids, `r toString(cc)`; however, Asteraceae `r toString(ff)`. Both morphotypes of Asteraceae pollen `r toString(dd)`, and Ericaceae `r toString(ee)`.

@@

par(pty="s")
plot(x = pol_p$countProp,  pol_p$adjSeqProp, 
      xlim = c(0,1), ylim = c(0,1), axes = F, main = 'Correlation of Proportion Counted\nGrains and Sequence Reads',
     xlab = 'Grains Counted', ylab = 'Classified Sequences',
     pch=18, slty = 3, gap = 0.025)
axis(1)
axis(2)
abline(model)
text(0.3, 0.9, paste0("tau = ", round(ct[["estimate"]][["tau"]], 3)))
text(0.3, 0.8, "p < 0.001")

rm( newdat)
```

@@

Layer   LM                     Description                              Source
----   ----      ------------------------------------------    -------------------------
1.      N              Mean annual cloudiness - MODIS              Wilson et al. 2016
2.      Y            Cloudiness seasonality 1 - MODIS              Wilson et al. 2016 
3.      N            Cloudiness seasonality 2 - MODIS              Wilson et al. 2016 
4.      Y            Cloudiness seasonality 3 - MODIS              Wilson et al. 2016
5.      N           Beginning of the frost-free period                 Wang et al.
6.      N              Climatic moisture deficit                       Wang et al.
7.      N              Degree-days above 5C from                       Wang et al.
8.      N              Mean annual precipitation                       Wang et al.
9.      Y           Mean annual precipitation as snow                  Wang et al.
10.     Y               Temperature seasonality                        Wang et al.
11.     Y         2015 Percent Grass/Herbaceous cover - MODIS           (MOD44B)
12.     Y          2015 Percent Tree cover from Landsat 7/8              (GLCF)
13.     Y         Soil probability of bedrock (R Horizon)               SoilGrids
14.     N           Soil organic carbon (Tonnes / ha)                   SoilGrids
15.     N                Surface soil pH in H~2~O                       SoilGrids
16.     Y               Surface soil percent sand                       SoilGrids
17.     Y                     Soil USDA class                           SoilGrids
18.     N                   Topographic elevation                      EarthEnv DEM
19.     Y             Topographic elevation, moving window.            EarthEnv DEM
20.     Y               Topographic percent slope                      EarthEnv DEM 
21.     Y               Topographic wetness index                      EarthEnv DEM 
22.     Y                  Topographic aspect                          EarthEnv DEM
23.     Y         Annual potential solar radiation computed               r.sun
24.     N        Estimated actual (w/-cloud) solar radiation      r.sun / Wilson et al. 2016
25.     Y        Log-transformed distance to surface water      Global Surface Water Explorer
26.     Y                Percent surface water                  Global Surface Water Explorer 
----   ----     -------------------------------------------    ------------------------------

>>>>

--- 
title: Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging Patterns of Queen Bumble Bees
author:  |
    |  Sophie Taddeo$^1$$^,$$^2$, Paul J. CaraDonna$^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant$^1$$^,$$^2$  
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    |  $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, Colorado 81224, USA
    |  $^4$ Australia post fire project, ----, ----, ----, AU  
    |  Reed Clark Benkendorf$^1$$^,$$^2$^[Correspondence: reedbenkendorf2021@u.northwestern.edu], Jane E. Ogilive$^3$$^,$$^4$, Emily J. Woodworth$^1$$^,$$^2$,
abstract:  |  
  |
  1) DNA Barcoding has been successful for the rapid analysis of ecological assemblages. Although barcoding in the plant kingdom barcoding has been more difficult than others, and hence may begin to lag behind other kingdoms. 
  2) Here we test the utilization of Angiosperms 353 probes to barcode plant species found in pollen loads collected from Queen Bumble Bees. 
  3) To verify the accuracy for this barcoding system we compared the data to museum species, observation studies, and species distribution modelling to identify likely candidate species. 
  4) By utilizing Species distribution modelling we allow users to create a regionally appropriate sequence databases which may use increase the alignment algorithms minimizing need for large computational power, and run time. 
  5) We show that the Angiosperms 353 probes, which are currently being used in the largest ever plant systematic endeavor, offers significant promise to metagenomic approaches. 
  6) Understanding plants in ecological contexts and understandings of their synecology.
keywords: |
  DNA baroding, metagenomics, hyb-seq, Angiosperms353, plant metagenomics, bombus queens, subalpine, species distribution model
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

pmorph <- read.csv('../data/morpho_pollen.csv') 
blst <- read.csv('../data/Post_Classified_BLAST_sqs.csv') %>% 
  rename(sample.id = Sample, Family = FAMILY, Genus = GENUS) %>% 
  group_by(sample.id) %>% 
  mutate(Prcnt_Seqs_Class = sum(Prcnt_seqs)) %>% 
  slice_max( n = 10, with_ties = F, order_by = Prcnt_seqs)
blst_seqs <- distinct(blst, sample.id)

@@

@@

aa <- prcnt_detect(mertensia_detected, Lprop = 0.01, Hprop = 1)
bb <- prcnt_detect(aster1_detected, Lprop = 0.001, Hprop = 0.01)
cc <- prcnt_detect(aster2_detected, Lprop = 0.001, Hprop = 0.01)
dd <- prcnt_detect(aster_detected, Lprop = 0.01, Hprop = 1)
ff <- prcnt_detect(aster_detected, Lprop = 0.001, Hprop = 0.01)

# Semi-quantitative accord between sequence % and Grain Count
palyn_cnt <- bind_rows(aster1_samples, aster2_samples, mertensia_samples) %>% 
  select(morphotype, countProp = Prop)

pollen_props <- bind_rows(blst_aster1, blst_aster2, blst_mertensia) %>% 
  select(morphotype, sequenceProp = Prcnt_seqs, Seqs_pr_taxon, Total_seqs, Prcnt_Seqs_Class) %>% 
  left_join(., palyn_cnt, by = c('sample.id', 'morphotype')) %>% 
  drop_na()

rm(mertensia_samples, blst_aster, blst_aster1,
   blst_aster2, aster1_samples,  aster2_samples, a, b, c, d, palyn_cnt,
   blst_mertensia, aster_samples)
```

@@

knitr::kable(pp_seqs_t, align = "c", 
             col.names = c('Condition', 'No. Class.', 'Prcnt. Class.', 'Total Seqs', 'Rank'),
             caption = "Post classification of Sequences via Taxonomy and Ecology,\ntop 15 most abundant reads") %>% 
  kableExtra::kable_styling(
      latex_options = "striped",
      stripe_color = "gray!10")

@@

To determine at which level species in pollen loads could be detected the results of light microscopy were compared to the molecular results. The pollen samples contained three morphotypes which could readily be identified via microscopy. Two of these mapped to the clades (Boraginaceae & Heliantheae Alliance), and one to a Asteraceae less Heliantheae. Boraginaceae grains `r toString(aa)`.  Asteraceae type 1, non-helianthoids, `r toString(bb)` Asteraceae type 2, Helianthoids, `r toString(cc)`; however, Asteraceae `r toString(ff)`. Both morphotypes of Asteraceae pollen `r toString(dd)`.

@@

# par(pty="s")
# plot(x = pol_p$countProp,  pol_p$adjSeqProp, 
#       xlim = c(0,1), ylim = c(0,1), axes = F, main = 'Correlation of Proportion Counted\nGrains and Sequence Reads',
#      xlab = 'Grains Counted', ylab = 'Classified Sequences',
#      pch=18, slty = 3, gap = 0.025)
# axis(1)
# axis(2)
# abline(model)
# text(0.3, 0.9, paste0("tau = ", round(ct[["estimate"]][["tau"]], 3)))
#text(0.3, 0.8, "p < 0.001")

# rm( newdat)
```

@@

Layer   LM                     Description                              Source
----   ----      ------------------------------------------    -------------------------
1.      N              Mean annual cloudiness - MODIS              Wilson et al. 2016
2.      Y            Cloudiness seasonality 1 - MODIS              Wilson et al. 2016 
3.      N            Cloudiness seasonality 2 - MODIS              Wilson et al. 2016 
4.      Y            Cloudiness seasonality 3 - MODIS              Wilson et al. 2016
5.      N           Beginning of the frost-free period                 Wang et al.
6.      N              Climatic moisture deficit                       Wang et al.
7.      N               Degree-days above 5C                           Wang et al.
8.      N              Mean annual precipitation                       Wang et al.
9.      Y           Mean annual precipitation as snow                  Wang et al.
10.     Y               Temperature seasonality                        Wang et al.
11.     Y         2015 Percent Grass/Herbaceous cover - MODIS           (MOD44B)
12.     Y          2015 Percent Tree cover from Landsat 7/8              (GLCF)
13.     Y         Soil probability of bedrock (R Horizon)               SoilGrids
14.     N           Soil organic carbon (Tonnes / ha)                   SoilGrids
15.     N                Surface soil pH in H~2~O                       SoilGrids
16.     Y               Surface soil percent sand                       SoilGrids
17.     Y                     Soil USDA class                           SoilGrids
18.     N                   Topographic elevation                      EarthEnv DEM
19.     Y           Topographic elevation, moving window.              EarthEnv DEM
20.     Y               Topographic percent slope                      EarthEnv DEM 
21.     Y               Topographic wetness index                      EarthEnv DEM 
22.     Y                  Topographic aspect                          EarthEnv DEM
23.     Y         Annual potential solar radiation computed               r.sun
24.     N        Estimated actual (w/-cloud) solar radiation      r.sun / Wilson et al. 2016
25.     Y        Log-transformed distance to surface water      Global Surface Water Explorer
26.     Y                Percent surface water                  Global Surface Water Explorer 
----   ----     -------------------------------------------    ------------------------------

### 2023-02-15 script/pollen_countr3.Rmd
new results in, lower detection?

pollen_df_new.1 <- pollen_df_new.1 %>%
  group_by(sample.id) %>% # form groups of each microscope slide. 
  mutate(count = as.numeric(count)) %>% # ensure that the values in this columns are represented as numbers to the computer
  mutate(count = replace_na(count, 0)) %>% # remember in our sample design NA's are 0's
  filter(transect >= 4, transect <= 17) %>%  # remove the 1st and 20th transects which we stopped counting. 
  mutate(group_sum = sum(count)) %>% # see which slides have been counted...
  drop_na(morphotype) %>% 
  filter(group_sum >= 1) %>% # find only the slides which we have counted pollen grains in .
  mutate(morphotype = str_replace(morphotype, "CAN\'T_DISTINGUISH", "UNKNOWN")) %>% 
  mutate(morphotype = str_replace(morphotype, "INDISTINGUISHABLE_CLUMP", "UNKNOWN")) %>% 
  mutate(morphotype = str_replace(morphotype, "NER", "UNKNOWN")) %>% 
  mutate(morphotype = str_replace(morphotype, "DEHYDRATED", "UNKNOWN")) %>%

  mutate(morphotype = str_replace(morphotype, "SALIX", "D")) %>% 
  mutate(morphotype = str_replace(morphotype, "C\\_\\(LINUM\\)", "LINUM")) %>% 
  mutate(morphotype = str_replace_all(morphotype, " ", "_")) %>% # replace white spaces 
  mutate(morphotype = str_to_upper(morphotype)) %>% # R is case sensitive
  mutate(morphotype = str_remove_all(morphotype, "\\_\\(11B\\)")) %>% # while good to know, unknown will be it's own class. 
  mutate(morphotype = if_else(str_detect(morphotype, "\\?"), "UNKNOWN", morphotype)) %>% # make all unknown observations the same format. 
  mutate(morphotype = str_replace(morphotype, "UNK$", "UNKNOWN")) %>% # make all unknown observations the same format. 
  mutate(morphotype = str_replace_all(morphotype, "UNKNWOWN", "UNKNOWN")) %>% 
  filter(!morphotype %in% c("NO_FJ", "LOW_FJ", "LITTLE_FJ", "LESS_FJ")) %>%  # we cannot say we counted or scored these cells. 
  filter(!sample.id %in% c(28, 29, 59))  %>%  # THIS SLIDES ARE PURELY OF A SINGLE TAXON, AND CANNOT BE ANALYZED. 
  filter(!morphotype %in% c(4,8)) %>% 
  mutate(plot.id = rleid(sample.id, transect, cell))

@@

write.csv(pollen_results, '/hdd/yun master/MEE_manuscript/data/morpho_pollen.csv', row.names = F)
```

@@

rich_rare <- data.table::rbindlist(lapply(to_rare[c(1:41,43:44,46:60)], raw_vegan), idcol = T) # a couple still not working, 
rich_rare <- rich_rare %>% rename(sample.id = '.id') %>% 
  mutate(sample.id = as.numeric(sample.id))

@@

#   Samples 45 and 48 missing due to only a single morphotype being present per slide. 
ggsave(
   "../graphics/plots/species_richness_rarefaction.png",
   plot = last_plot(),
   scale = 1,
   width = 20,
   height =28,
   units =  "cm",
   dpi = 300,
)

>>>>

pollen_df_new.1 <- pollen_df_new.1 %>%
  group_by(sample.id) %>% # form groups of each microscope slide. 
  mutate(count = as.numeric(count)) %>% # ensure that the values in this columns are represented as numbers to the computer
  mutate(count = replace_na(count, 0)) %>% # remember in our sample design NA's are 0's
  mutate(group_sum = sum(count)) %>% # see which slides have been counted...
  drop_na(morphotype) %>% 
  filter(group_sum >= 1) %>% # find only the slides which we have counted pollen grains in .
  mutate(
    morphotype = str_to_upper(morphotype),
    morphotype = str_replace(morphotype, "CAN\'T_DISTINGUISH", "UNKNOWN"),
    morphotype = str_replace(morphotype, "INDISTINGUISHABLE_CLUMP", "UNKNOWN"),
    morphotype = str_replace(morphotype, "NER", "UNKNOWN"),
    morphotype = str_replace(morphotype, "DEHYDRATED", "UNKNOWN"),
    morphotype = str_replace(morphotype, 'UNKNOWN_[(]5B[)]', 'UNKNOWN'),
    morphotype = str_replace(morphotype, '5B', 'UNKNOWN'),
    morphotype = str_replace(morphotype, 'ASTERACEAE-2', 'ASTERACEAE_2'),
    morphotype = str_replace(morphotype, "SALIX", "D"),
    morphotype = str_replace(morphotype, 'C4', 'C'),
    morphotype = str_replace(morphotype, 'OENOTHERA', 'ONAGRACEAE'),
    morphotype = str_replace(morphotype, 'UNKNOWN_[(]D[)]', 'D'),
    morphotype = str_replace(morphotype, 'OTHER_B', 'B'),
    morphotype = str_replace(morphotype, 'OTHER_C', 'C'),
    morphotype = str_replace(morphotype, "C\\_\\(LINUM\\)", "LINUM"),
    morphotype = str_replace(morphotype, 'UNKNOWN_[(]RANUNCULUS[)]', 'UNKNOWN'),
    morphotype = str_replace(morphotype, 'UNKNOWN_[(]11B[/]D[)]', 'UNKNOWN'),
    morphotype = str_replace(morphotype, 'UNKNOWN_[(]4_APERTURES[)]', 'UNKNOWN'),
    morphotype = str_replace_all(morphotype, " ", "_"),
    morphotype = str_remove(morphotype, "\\_\\(11B\\)"),
    morphotype = str_replace_all(morphotype, "UNKNWOWN", "UNKNOWN")) %>% 
  filter(!morphotype %in% c("NO_FJ", "LOW_FJ", "LITTLE_FJ", "LESS_FJ", 'LILTTLE_FJ')) %>%  # we cannot say we counted or scored these cells. 
  filter(!morphotype %in% c(4,8, 11)) %>% 
  mutate(plot.id = rleid(sample.id, transect, cell))

@@

write.csv(pollen_results, '../data/morpho_pollen.csv', row.names = F)
```

@@

#[c(1:41,43:44,46:60)]
# 45 does not work
u <- lapply(lapply(to_rare, rowSums), sum)

to_rare <- to_rare[u > 1]
rich_rare <- data.table::rbindlist(lapply(to_rare, raw_vegan), idcol = T)

rich_rare <- rich_rare %>% rename(sample.id = '.id') %>% 
  mutate(sample.id = as.numeric(sample.id))

@@

#   Samples 45 iz missing due to only a single morphotype being present per slide. 
ggsave(
   "../graphics/plots/species_richness_rarefaction.png",
   plot = last_plot(),
   scale = 1,
   width = 20,
   height =28,
   units =  "cm",
   dpi = 300,
)
